Reference Tracking Strategies of Deaf Adult Signers in Turkish Sign Language

### 2023-05-02 revised_manuscript.tex
model results added
Added the model results to the appendix.

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Reference Tracking Strategies of Deaf Adult Signers in Turkish Sign Language},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% Pandoc header
\usepackage{floatrow}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{float}
\floatplacement{figure}{H}
\floatplacement{table}{H}
\floatsetup[figure]{capposition=top}
\floatsetup[table]{capposition=top}
\usepackage{times}
\usepackage{smartdiagram}
\usepackage{ragged2e}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{./sty/langsci-gb4e}
\usepackage{caption}
\usepackage{multicol}
\setlength{\columnsep}{-2cm}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Reference Tracking Strategies of Deaf Adult Signers in Turkish
Sign Language}

  \begin{abstract}
  We investigated the reference tracking strategies among deaf adults
  with signing deaf parents (DoD) and adult deaf signers with
  non-signing hearing parents or caregivers (DoH) who had late exposure
  to Turkish Sign Language (TİD). Consistent with the theories of
  saliency and referential accessibility, regardless of their acquisiton
  groups or parental hearing status, signers mainly used nominals and
  extension classifiers for introducing referents. To maintain a
  referent from the previous clause, the signers used zero anaphora
  (e.g., constructed action, agreement and plain verbs). For topic
  shifts or re-introduced contexts, nominals and pronouns were chiefly
  favored although we observed very little pronominal use. As for the
  effect of native acquisition from deaf caregivers, we only report
  limited over-redundancy for DoH signers who used zero anaphora less
  compared to DoD especially for introduced and maintained contexts. We
  conclude that DoH signers are still able to achieve native-like
  competency in terms of reference tracking in simple narratives but DoD
  signers utilize the spatial affordances of the visual-spatial modality
  better than DoH signers only to a certain extent.
  \end{abstract}
   \begin{keyword} Referential accessibility, signed narratives,
pragmatic competence, delayed language acquisition, Turkish Sign
Language\end{keyword}
 \end{frontmatter}

@@

Narratives are extensive discourses that form an important part of human
communication. In order to create a systematic and coherent organization
of experiences, narrators must choose appropriate linguistic forms to
refer to entities (Ariel, 1990; Givon, 1983; Gundel et al., 1993).
Certain forms such as nominal expressions are commonly used for the
first mention of a referent in a discourse, and other forms such as
pronouns and zero anaphora may track previously mentioned entities which
could either be maintained across multiple clauses or re-introduced back
into the discourse. The selection of these referring expressions
(henceforth REs) is contingent on cognitive abilities, such that
individuals must rely on the activation of the working memory to keep in
mind previously referred expressions and long strings of information
(Bamberg, 1997; Morgan, 2005). Pragmatic abilities too play a pivotal
role in the tracking of referents. The form with the least amount of
redundant information must be selected based on the accessibility levels
of the referent (i.e., how active a referent is in the addressee's mind
given a point in discourse) (Ahn, 2019; Ariel, 1990). Because of these
cognitive and pragmatic necessities, it is not surprising to observe
that producing a coherent narrative is one of the latest milestones in
language acquisition among children (Hickmann et al., 1996). As can be
expected, there is a plethora of research examining how narrative
competence is achieved in both monolingual and bilingual acquisition
contexts among children and adults. Although narrative competence and
referent selection have been studied extensively for typically
developing speakers (Aksu-Koç and Nicolopoulou, 2015; Hickmann et al.,
1996; Williams, 1988), deaf individuals who acquired a sign language in
atypical learning contexts (i.e., deaf signers with delayed exposure to
a sign language) have remained mostly understudied with an exception of
a handful of studies (Becker, 2009; Cormier et al., 2013; Gür, 2018).
Nevertheless, it is paramount that we investigate and describe the
effects of late exposure to a sign language among deaf adults growing up
in a signing vs.~non-signing environment on narrative competence more in
depth. This will help us understand whether the lack of systematic and
frequent language exposure in the early years of life affect how deaf
signers produce a narrative in their adulthood. Given this, the present
study will focus on two groups of signers and examine their reference
tracking strategies in signed narratives: (i) deaf of deaf (DoD) adult
signers (i.e., signers who have deaf caregivers and natively acquired a
sign language from birth) and (ii) deaf of hearing (DoH) adult signers
(i.e., signers who have hearing non-signing caregivers and state to have
started acquiring a sign language later in life).

Sign languages are the natural languages of the Deaf communities all
around the world. They are independent languages, which do not derive
from majority spoken language(s) but emerge naturally when deaf people
form a community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and
has been the officially recognized language of the Deaf community in
Turkey since 2005 (see Nuhbalaoglu, 2018 for a brief review of the
historical development of the language). Deaf or hearing children
acquiring a SL within a family where there is one or more deaf
caregivers display the same developmental milestones as hearing children
acquiring a spoken language in a hearing family (Chamberlain et al.,
1999; Meier and Newport, 1990). However, reports indicate that over 90\%
of all deaf individuals are born into hearing families (DoH) who do not
know how to sign (Woll, 2013). This means that the auditory modality
used in the household is not accessible to many deaf children. That is
why the age of acquisition of a first functional language varies greatly
among deaf children, with many of whom being deprived of their first
languages typically until after age 7 when they start formal education
in a deaf school, especially in the context of Turkey (Sari, 2005).
Despite this, many of the early intervention programs for late signing
children prioritize spoken language development and discourage the use
of a sign language (Kemaloğlu and Kemaloğlu, 2012). The Ministry of
National Education (2020) identifies 72 deaf schools in Turkey for the
K-12 level. Even when deaf children with hearing parents start their
formal education in one of these schools, they do not receive proper
instruction in TİD since the teachers themselves are mainly hearing and
only have a very rudimentary knowledge of a sign language (Ilkbasaran,
2015; Sari, 2005). In some cases, it has been argued that late exposure
to a sign language among certain DoH signers has negative effects on
some cognitive and linguistic measures. To name a few, delayed exposure
to a first functional language is reported to negatively affect deaf
children and adults' morphosyntactic production and processing
(Boudreault and Mayberry, 2006; Cheng and Mayberry, 2021; Cormier et
al., 2012; Karadöller et al., 2017; Kayabaşı and Gökgöz, 2022; Newport,
1990) and narrative abilities (e.g., Becker, 2009). In our overall
research, we are concerned with the effects of this late exposure in the
ultimate adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy (Gundel et al., 1993).
According to this model, given or familiar information is expressed in
more attenuated forms such as pronouns and zero anaphora (i.e., there is
no explicit mention of the referent) since the mental status of the
expressed referent is already at the center of attention, as indicated
by the present contextual information (Chafe, 1976). The Givenness
Hierarchy is akin to other previously proposed models of reference
control within the topicality and saliency paradigm such as the
Accessibility Hierarchy (Ariel, 1990), Topic Continuity (Givon, 1983),
and Familiarity Scale (Prince, 1981). Gundel et al. (1993) argued that
the distribution of forms is not random but abides by the Gricean Maxim
of Quantity (Grice, 1975). That is to say, individuals must make the
most informative and succinct choice possible to retain economy of form.
When multiple expressions are available for the same referent as in an
utterance like (\ref{ex:1}), phonetically simplest form is selected with
the least redundant information based on an economy principle. In the
utterance, REs for ``Eddie'' are given in bold. All three REs (nominal,
pronominal, and zero anaphora) refer to the same entity; however, the
quantity of marking decreases as the referent's accessibility increases.
On this ground, for highly accessible referents (i.e., those that are
highly active in the mind), linguistic forms with the least phonetic
content are selected for the least amount of processing effort.

@@

It is also possible to categorize REs into three according to their
discourse status (Gullberg, 2006). A referent can be introduced or
mentioned for the first time, maintained, that is, continued across two
or more clauses, or re-introduced into the discourse following a topic
shift. As a result, interlocutors mark some REs differently than others
based on the principles of reference accessibility. For spoken
languages, this observation is well established (Ahrenholz, 2005; Ariel,
1990; Azar and Özyürek, 2015; Chafe, 1976; Debreslioska et al., 2013;
Givon, 1983; Gundel et al., 1993; Hickmann et al., 1996). These research
indicate that the introduction and re-introduction of a referent are
often accomplished by full noun phrases which can be commonly modified
by an adjective or demonstrative. Maintenance, on the other hand, is
typically marked with pronominals and zero anaphora since the continued
referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives (Ahlgren and Bergman, 1994;
Czubek, 2017; Ferrara et al., 2022; Frederiksen and Mayberry, 2016;
Garcia et al., 2018; Garcia and Sallandre, 2014; Hodge et al., 2019;
Morgan, 2005; Perniss and Özyürek, 2015; Swabey, 2002). Building on the
work by Gundel et al. (1993), Swabey (2002) demonstrated that, like
other languages such as Russian, Chinese, and Japanese, American Sign
Language (ASL) signers use bare nouns for introduced and re-introduced
discourse contexts whereas zero anaphora is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and thus used more economic forms of reference
tracking. Similar findings have been reported for German Sign Language
(Deutsche Gebärdensprache - DGS) (Perniss and Özyürek, 2015) and British
Sign Language (BSL) (Morgan, 2005). It is important to note, however,
that this research either collapsed DoD and DoH signers into a single
group and labeled them as ``native-like'' or explicitly compared native
deaf signers with late deaf signers (with delayed exposure to a sign
language) in terms of narrative competence. Most recently, Ferrara et
al. (2022) examined corpora including video-recorded narratives from
five different sign languages including Irish Sign Language, Finnish
Sign Language, Norwegian Sign Language and Swedish Sign Language and
investigated how these languages established and maintained referents.
Like other observations, this research showed that, across each of these
languages, bare nouns were used to introduce new referents and zero
anaphora or more reduced forms were preferred for referent maintenance.

@@

The linguistic inventory for discourse referents can include nominal
constructions (bare nouns, fingerspelled nouns, and modified nouns),
null or zero
anaphors\footnote{Zero anaphora does not refer to the lack of a referent. It only alludes to the non-presence of a more explicit anaphoric form for the referent such as a nominal, pronominal, or classifiers.}
(predicated by a plain verb, an agreement verb, or constructed action),
classifier predicates (whole entity classifiers, body part classifiers,
handling classifiers, and extension classifiers), and overt pronouns
(pronominal index or IX) (Frederiksen and Mayberry, 2016). This interim
section describes such forms attested for sign languages under four main
headings (nominals, pronouns, classifiers, and zero anaphora).

@@

In TİD and many other sign languages, it is possible to use bare nouns
as in \ref{ex:2a} as well as fingerspelled nouns as in \ref{ex:2b}.
Fingerspelling refers to the use of the manual alphabet to spell all or
part of the letters of a word borrowed from a surrounding written
language (Kubuş, 2008). Nouns can also be modified by a prenominal or
postnominal extended index finger pointing sign, glossed as IX,
functioning as a determiner or a demonstrative as in \ref{ex:2c}
(Maclaughlin, 1997) or by a modifying classifier within the same NP as
in \ref{ex:2d} (Frederiksen and Mayberry,
2016).\footnote{Spoken languages use pronouns, determines and demonstratives where sign languages use a pointing sign, usually glossed as IX, which assumes similar functions (Lillo-Martin and Klima, 1990). In addition, the pointing sign is used to referentially anchor a referent to a location in signing space.}

@@

In addition to being a modifier of a noun, the IX sign can also be
pronominal by referring to established nominals through pointing to a
previously assigned abstract or physical locus in the signing space when
re-introducing a topic in discourse (Emmorey, 1996). This function has
also been identified for TİD (Sevinç, 2006). The example in \ref{ex:3}
illustrates this spatially-anchored pronominal use.

@@

These pronominal signs are rarely used for the first mention of an
entity (Perniss and Özyürek, 2015). Other researchers consider pointing
signs without noun phrases as examples of demonstratives and not
personal pronouns (Koulidobrova and Lillo-Martin, 2016). It is also
important to mention here that in some spoken languages (e.g., German
and Dutch), demonstratives can also refer to persons (Bosch and Umbach,
2007) and behave similarly to personal pronouns. However, for the
purposes of the present study, we will refer to IX signs as pronominal
IX and particularly examine their non-deictic or personal pronoun
realizations as a possible discourse strategy to mainly mark topic shift
(re-introduction) among DoD and DoH signers of TİD.

@@

In sign languages, a verbal classifier morpheme is morpho-phonologically
encoded by a handshape. This handshape is articulated co-temporally with
a movement that encodes the verbal meaning (Benedicto and Brentari,
2004; Supalla, 1990). Semantically, the classifiers denote sets of
objects based on criteria such as animacy, size, shape, body-part,
texture and sometimes function (Talmy, 2003). The classifier articulated
simultaneously with a movement is called a classifier predicate and the
sentence in which such a predicate is used is usually called a
classifier construction in sign language research. As in most sign
languages hitherto studied, classifier constructions are also commonly
found in TİD (see Dikyuva et al., 2015 for a review). Following the
classification of (Engberg-Pedersen, 1993), in TİD, we distinguish
between four types of classifier handshapes. Whole entity classifiers
(WECL) represent the entirety of a referent (e.g.~an entire flat
horizontal entity, or an upright entity) (e.g., \ref{ex:4a}) whereas
body part classifiers (BPCL) represent a specific part of an animate
entity (e.g., limbs, head, mouth, tongue, or teeth, etc.)
(e.g.,\ref{ex:4b}). Handling classifiers (HCL) occur with transitive
verbal roots and represent both the size, shape and surface of an object
and how that object is handled or manipulated by an agent (e.g.,
\ref{ex:4c}). Extension classifiers (ExtCL), on the other hand,
represent the size and shape of an entity by following a trajectory to
trace its shape (e.g., \ref{ex:4d}). We exemplify these handshapes with
the discourse status of the referent they track and a description of the
previous discourse when available.

@@

In both communication modalities (visual and auditory), children start
to produce more complex and coherent stories with appropriate and
informative linguistic forms to indicate anaphoric or cataphoric
references after age 10 (Hickmann et al., 1996; Rathmann et al., 2007).
Prior to this age, both signing and speaking children fail to refer to
entities consistently since their pragmatic abilities are yet to
develop. The manipulation of linguistic forms for discourse functions to
signal topic continuance or shift is difficult to master until the
beginning of formal education. These findings are also comparable across
different sign languages including BSL (Morgan, 2005), ASL (Loew, 1984),
and DGS (Becker, 2009), pointing out to the fact that the mastery of
reference tracking in narratives is one of the most protracted
achievements in the acquisition of signed discourse.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Reference Tracking Strategies of Deaf Adult Signers in Turkish Sign Language},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{floatrow}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{float}
\floatplacement{figure}{H}
\floatplacement{table}{H}
\floatsetup[figure]{capposition=top}
\floatsetup[table]{capposition=top}
\usepackage{times}
\usepackage{smartdiagram}
\usepackage{ragged2e}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{langsci-gb4e}
\usepackage{caption}
\usepackage{multicol}
\setlength{\columnsep}{-2cm}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\begin{frontmatter}

  \title{Reference Tracking Strategies of Deaf Adult Signers in Turkish
Sign Language}
      \cortext[cor1]{Corresponding author}

  \begin{abstract}
  We investigated the reference tracking strategies among deaf adults
  with signing deaf parents (DoD) and adult deaf signers with
  non-signing hearing parents or caregivers (DoH) who had late exposure
  to Turkish Sign Language (TİD). Consistent with the theories of
  saliency and referential accessibility, regardless of their acquisiton
  groups or parental hearing status, signers mainly used nominals and
  extension classifiers for introducing referents. To maintain a
  referent from the previous clause, the signers used zero anaphora
  (e.g., constructed action, agreement and plain verbs). For topic
  shifts or re-introduced contexts, nominals and pronouns were chiefly
  favored although we observed very little pronominal use. As for the
  effect of native acquisition from deaf caregivers, we only report
  limited over-redundancy for DoH signers who used zero anaphora less
  compared to DoD especially for introduced and maintained contexts. We
  conclude that DoH signers are still able to achieve native-like
  competency in terms of reference tracking in simple narratives but DoD
  signers utilize the spatial affordances of the visual-spatial modality
  better than DoH signers only to a certain extent.
  \end{abstract}
    \begin{keyword}
    Referential accessibility, signed narratives, pragmatic competence,
delayed language acquisition, Turkish Sign Language \sep 
    Referential accessibility, signed narratives, pragmatic competence,
delayed language acquisition, Turkish Sign Language
  \end{keyword}

 \end{frontmatter}

@@

Narratives are extensive discourses that form an important part of human
communication. In order to create a systematic and coherent organization
of experiences, narrators must choose appropriate linguistic forms to
refer to entities \citep{ariel1990, givon1983, gundel1993}. Certain
forms such as nominal expressions are commonly used for the first
mention of a referent in a discourse, and other forms such as pronouns
and zero anaphora may track previously mentioned entities which could
either be maintained across multiple clauses or re-introduced back into
the discourse. The selection of these referring expressions (henceforth
REs) is contingent on cognitive abilities, such that individuals must
rely on the activation of the working memory to keep in mind previously
referred expressions and long strings of information
\citep{morgan2005, bamberg1997}. Pragmatic abilities too play a pivotal
role in the tracking of referents. The form with the least amount of
redundant information must be selected based on the accessibility levels
of the referent (i.e., how active a referent is in the addressee's mind
given a point in discourse) \citep{ahn2019, ariel1990}. Because of these
cognitive and pragmatic necessities, it is not surprising to observe
that producing a coherent narrative is one of the latest milestones in
language acquisition among children \citep{hickmann1996}. As can be
expected, there is a plethora of research examining how narrative
competence is achieved in both monolingual and bilingual acquisition
contexts among children and adults. Although narrative competence and
referent selection have been studied extensively for typically
developing speakers \citep{aksu-koc2015, hickmann1996, williams1988},
deaf individuals who acquired a sign language in atypical learning
contexts (i.e., deaf signers with delayed exposure to a sign language)
have remained mostly understudied with an exception of a handful of
studies \citep{cormier2013, gur2018, becker2009}. Nevertheless, it is
paramount that we investigate and describe the effects of late exposure
to a sign language among deaf adults growing up in a signing
vs.~non-signing environment on narrative competence more in depth. This
will help us understand whether the lack of systematic and frequent
language exposure in the early years of life affect how deaf signers
produce a narrative in their adulthood. Given this, the present study
will focus on two groups of signers and examine their reference tracking
strategies in signed narratives: (i) deaf of deaf (DoD) adult signers
(i.e., signers who have deaf caregivers and natively acquired a sign
language from birth) and (ii) deaf of hearing (DoH) adult signers (i.e.,
signers who have hearing non-signing caregivers and state to have
started acquiring a sign language later in life).

Sign languages are the natural languages of the Deaf communities all
around the world. They are independent languages, which do not derive
from majority spoken language(s) but emerge naturally when deaf people
form a community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and
has been the officially recognized language of the Deaf community in
Turkey since 2005 \citep[see][ for a brief review of the historical
development of the language]{nuhbalaoglu2018}. Deaf or hearing children
acquiring a SL within a family where there is one or more deaf
caregivers display the same developmental milestones as hearing children
acquiring a spoken language in a hearing family
\citep{chamberlain_language_1999, meier1990}. However, reports indicate
that over 90\% of all deaf individuals are born into hearing families
(DoH) who do not know how to sign \citep{woll2013}. This means that the
auditory modality used in the household is not accessible to many deaf
children. That is why the age of acquisition of a first functional
language varies greatly among deaf children, with many of whom being
deprived of their first languages typically until after age 7 when they
start formal education in a deaf school, especially in the context of
Turkey \citep{sari_analysis_2005}. Despite this, many of the early
intervention programs for late signing children prioritize spoken
language development and discourage the use of a sign language
\citep{kemaloglu2012}. The \citet{ministryofnationaleducation2020}
identifies 72 deaf schools in Turkey for the K-12 level. Even when deaf
children with hearing parents start their formal education in one of
these schools, they do not receive proper instruction in TİD since the
teachers themselves are mainly hearing and only have a very rudimentary
knowledge of a sign language \citep{ilkbasaran2015, sari_analysis_2005}.
In some cases, it has been argued that late exposure to a sign language
among certain DoH signers has negative effects on some cognitive and
linguistic measures. To name a few, delayed exposure to a first
functional language is reported to negatively affect deaf children and
adults' morphosyntactic production and processing
\citep{kayaba2022, boudreault2006, newport1990, cheng2021, karadoller2017, cormier2012}
and narrative abilities \citep[e.g.,][]{becker2009}. In our overall
research, we are concerned with the effects of this late exposure in the
ultimate adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy \citep{gundel1993}.
According to this model, given or familiar information is expressed in
more attenuated forms such as pronouns and zero anaphora (i.e., there is
no explicit mention of the referent) since the mental status of the
expressed referent is already at the center of attention, as indicated
by the present contextual information \citep{chafe1976}. The Givenness
Hierarchy is akin to other previously proposed models of reference
control within the topicality and saliency paradigm such as the
Accessibility Hierarchy \citep{ariel1990}, Topic Continuity
\citep{givon1983}, and Familiarity Scale \citep{prince1981}.
\citet{gundel1993} argued that the distribution of forms is not random
but abides by the Gricean Maxim of Quantity \citep{grice1975}. That is
to say, individuals must make the most informative and succinct choice
possible to retain economy of form. When multiple expressions are
available for the same referent as in an utterance like (\ref{ex:1}),
phonetically simplest form is selected with the least redundant
information based on an economy principle. In the utterance, REs for
``Eddie'' are given in bold. All three REs (nominal, pronominal, and
zero anaphora) refer to the same entity; however, the quantity of
marking decreases as the referent's accessibility increases. On this
ground, for highly accessible referents (i.e., those that are highly
active in the mind), linguistic forms with the least phonetic content
are selected for the least amount of processing effort.

@@

It is also possible to categorize REs into three according to their
discourse status \citep{gullberg2006}. A referent can be introduced or
mentioned for the first time, maintained, that is, continued across two
or more clauses, or re-introduced into the discourse following a topic
shift. As a result, interlocutors mark some REs differently than others
based on the principles of reference accessibility. For spoken
languages, this observation is well established
\citep{ariel1990, givon1983, gundel1993, debreslioska2013, ahrenholz2005, chafe1976, hickmann1996, azar2015}.
These research indicate that the introduction and re-introduction of a
referent are often accomplished by full noun phrases which can be
commonly modified by an adjective or demonstrative. Maintenance, on the
other hand, is typically marked with pronominals and zero anaphora since
the continued referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives
\citep{czubek2017, ferrara2022, swabey2002, frederiksen2016, perniss2015, morgan2005, hodge2019, ahlgren1994, garcia_reference_2014, garcia2018impersonal}.
Building on the work by \citet{gundel1993}, \citet{swabey2002}
demonstrated that, like other languages such as Russian, Chinese, and
Japanese, American Sign Language (ASL) signers use bare nouns for
introduced and re-introduced discourse contexts whereas zero anaphora is
used for referents which are ``in-focus'' or very accessible to the
addressee. Overt REs like pronouns or nouns are disprefered for such
referents. \citet{swabey2002} concluded that ``native'' signers in her
study abided by the Gricean maxim of quantity and thus used more
economic forms of reference tracking. Similar findings have been
reported for German Sign Language (Deutsche Gebärdensprache - DGS)
\citep{perniss2015} and British Sign Language (BSL) \citep{morgan2005}.
It is important to note, however, that this research either collapsed
DoD and DoH signers into a single group and labeled them as
``native-like'' or explicitly compared native deaf signers with late
deaf signers (with delayed exposure to a sign language) in terms of
narrative competence. Most recently, \citet{ferrara2022} examined
corpora including video-recorded narratives from five different sign
languages including Irish Sign Language, Finnish Sign Language,
Norwegian Sign Language and Swedish Sign Language and investigated how
these languages established and maintained referents. Like other
observations, this research showed that, across each of these languages,
bare nouns were used to introduce new referents and zero anaphora or
more reduced forms were preferred for referent maintenance.

@@

The linguistic inventory for discourse referents can include nominal
constructions (bare nouns, fingerspelled nouns, and modified nouns),
null or zero
anaphors\footnote{Zero anaphora does not refer to the lack of a referent. It only alludes to the non-presence of a more explicit anaphoric form for the referent such as a nominal, pronominal, or classifiers.}
(predicated by a plain verb, an agreement verb, or constructed action),
classifier predicates (whole entity classifiers, body part classifiers,
handling classifiers, and extension classifiers), and overt pronouns
(pronominal index or IX) \citep{frederiksen2016}. This interim section
describes such forms attested for sign languages under four main
headings (nominals, pronouns, classifiers, and zero anaphora).

@@

In TİD and many other sign languages, it is possible to use bare nouns
as in \ref{ex:2a} as well as fingerspelled nouns as in \ref{ex:2b}.
Fingerspelling refers to the use of the manual alphabet to spell all or
part of the letters of a word borrowed from a surrounding written
language \citep{kubus2008}. Nouns can also be modified by a prenominal
or postnominal extended index finger pointing sign, glossed as IX,
functioning as a determiner or a demonstrative as in \ref{ex:2c}
\citep{maclaughlin1997} or by a modifying classifier within the same NP
as in \ref{ex:2d}
\citep{frederiksen2016}.\footnote{Spoken languages use pronouns, determines and demonstratives where sign languages use a pointing sign, usually glossed as IX, which assumes similar functions (Lillo-Martin and Klima, 1990). In addition, the pointing sign is used to referentially anchor a referent to a location in signing space.}

@@

In addition to being a modifier of a noun, the IX sign can also be
pronominal by referring to established nominals through pointing to a
previously assigned abstract or physical locus in the signing space when
re-introducing a topic in discourse \citep{emmorey1996}. This function
has also been identified for TİD \citep{sevinc2006}. The example in
\ref{ex:3} illustrates this spatially-anchored pronominal use.

@@

These pronominal signs are rarely used for the first mention of an
entity \citep{perniss2015}. Other researchers consider pointing signs
without noun phrases as examples of demonstratives and not personal
pronouns \citep{koulidobrova2016}. It is also important to mention here
that in some spoken languages (e.g., German and Dutch), demonstratives
can also refer to persons \citep{bosch2007} and behave similarly to
personal pronouns. However, for the purposes of the present study, we
will refer to IX signs as pronominal IX and particularly examine their
non-deictic or personal pronoun realizations as a possible discourse
strategy to mainly mark topic shift (re-introduction) among DoD and DoH
signers of TİD.

@@

In sign languages, a verbal classifier morpheme is morpho-phonologically
encoded by a handshape. This handshape is articulated co-temporally with
a movement that encodes the verbal meaning
\citep{supalla1990, benedicto2004}. Semantically, the classifiers denote
sets of objects based on criteria such as animacy, size, shape,
body-part, texture and sometimes function \citep{talmy2003}. The
classifier articulated simultaneously with a movement is called a
classifier predicate and the sentence in which such a predicate is used
is usually called a classifier construction in sign language research.
As in most sign languages hitherto studied, classifier constructions are
also commonly found in TİD \citep[see][ for a review]{dikyuva2015}.
Following the classification of \citep{engberg-pedersen1993}, in TİD, we
distinguish between four types of classifier handshapes. Whole entity
classifiers (WECL) represent the entirety of a referent (e.g.~an entire
flat horizontal entity, or an upright entity) (e.g., \ref{ex:4a})
whereas body part classifiers (BPCL) represent a specific part of an
animate entity (e.g., limbs, head, mouth, tongue, or teeth, etc.)
(e.g.,\ref{ex:4b}). Handling classifiers (HCL) occur with transitive
verbal roots and represent both the size, shape and surface of an object
and how that object is handled or manipulated by an agent (e.g.,
\ref{ex:4c}). Extension classifiers (ExtCL), on the other hand,
represent the size and shape of an entity by following a trajectory to
trace its shape (e.g., \ref{ex:4d}). We exemplify these handshapes with
the discourse status of the referent they track and a description of the
previous discourse when available.

@@

In both communication modalities (visual and auditory), children start
to produce more complex and coherent stories with appropriate and
informative linguistic forms to indicate anaphoric or cataphoric
references after age 10 \citep{hickmann1996, rathmann2007}. Prior to
this age, both signing and speaking children fail to refer to entities
consistently since their pragmatic abilities are yet to develop. The
manipulation of linguistic forms for discourse functions to signal topic
continuance or shift is difficult to master until the beginning of
formal education. These findings are also comparable across different
sign languages including BSL \citep{morgan2005}, ASL \citep{loew1984},
and DGS \citep{becker2009}, pointing out to the fact that the mastery of
reference tracking in narratives is one of the most protracted
achievements in the acquisition of signed discourse.
